---
title: "Michelson Speed-of-light Measurements"
author: "Daeyoung Kim"
date: 2023-01-30
output:
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
---

*Purpose*: When studying physical problems, there is an important distinction between *error* and *uncertainty*. The primary purpose of this challenge is to dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)
library(lintr)
url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure the speed of light. His approach was a development upon the method of Foucault[3], and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value `LIGHTSPEED_VACUUM`) because the meter is actually [*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

### __q1__ Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- df_michelson
df_q1 %>%
  group_by(Distinctness) %>%
  summarise(n = n(), MeanVelocity = mean(Velocity)) %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:
- Write your observations here!
  - Trials appear the most frequent to infrequent with descending order of 'Distinctness'.
  - The 'MeanVelocity' is also in descending magnitudes when you go from the most frequent to least. 
- Why might your table differ from Michelson's?
  - The table above still contains precision to the tenths whereas Michelson's seems to have been round to the nearest tens. 

The `Velocity` values in the dataset are the speed of light *in air*; Michelson
introduced a couple of adjustments to estimate the speed of light in a vacuum.
In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from
Michelson (1880), pg. 141). While the following isn't fully rigorous ($+92$ km/s
is based on the mean temperature), we'll simply apply this correction to all the
observations in the dataset.

### __q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <-
  df_michelson %>% 
  mutate(VelocityVacuum = Velocity + 92)
df_q2
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

### __q3__ Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
between(
  LIGHTSPEED_VACUUM, 
  LIGHTSPEED_MICHELSON - LIGHTSPEED_PM, 
  LIGHTSPEED_MICHELSON + LIGHTSPEED_PM
  )
error <- abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON)
uncertainty <- LIGHTSPEED_PM
difference <- error - uncertainty

error
uncertainty
difference
```

**Observations**:
- Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?
  - The estimate of the error was less than the true error.
- Make a quantitative comparison between Michelson's uncertainty and his error.
  - The uncertainty was about 1/3 of the magnitude of his actual error -- it was about 100 km/s off from the error. 

The following plot shows all of Michelson's data as a [control chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot is common in manufacturing, where it is used to help determine if a manufacturing process is under [statistical control](https://en.wikipedia.org/wiki/Statistical_process_control). Each dot is one of Michelson's measurements, and the grey line connects the mean taken for each day. The same plot also shows simulated data using a probability model. We'll get into statistics later in the course; for now, let's focus on understanding what real and simulated data tend to look like.

### __q4__ Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source ~ .) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:
Similarities
- The mean velocities fluctuate but both stays generally within the bounds from 300,000 to 299,900 km/s. 
- 
Differences
- The dots in each day are more scattered in the 'Real' plot compared to the 'Simulated' plot. 
- The mean velocities stay closer to the middle of the bounds in the 'Simulated' plot than the 'Real' plot.

### __q5__ You have access to a few other variables. Construct a few visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?
```{r}
df_q2 %>%
  group_by(Temp) %>%
  mutate(mean = mean(VelocityVacuum)) %>%
  ggplot(aes(x = Temp, y = VelocityVacuum)) +
  geom_point() +
  geom_smooth(
    mapping = aes(color = "Trendline of the Velocity"), 
    formula = y ~ x, 
    method = lm, 
    se = FALSE, 
    linewidth = 0.5
    ) +
  geom_hline(mapping = aes(
    yintercept = LIGHTSPEED_MICHELSON,
    color = "Michelson's Velocity of Light",
    ),
    linetype = "dashed",
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dotted"
  ) +
  geom_hline(
    mapping = aes(
      yintercept = LIGHTSPEED_VACUUM, 
      color = "Modern Velocity of Light"
      ), 
    linetype = "dashed",
    alpha = 0.4
  ) +
  scale_color_manual(
    values = c(
      "Michelson's Velocity of Light" = "brown",
      "Modern Velocity of Light" = "blue",
      "Trendline of the Velocity" = "red"
    )
  ) +
  labs(
    title = "Michelson's Velocity Measurements vs Temperature",
    caption = str_wrap("The plot shows the relationship between Michelson's 
                       measurements of velocity of light and temperature.",
                       80),
    y = "Experimental Velocity of Light (km/s)",
    x = "Temperature (°F)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5),
    legend.title = element_blank()
    )
  
```
```{r}
df_q2 %>% 
  ggplot(mapping = aes(x = Distinctness)) +
  geom_point(
    mapping = aes(y = VelocityVacuum)
  ) +
  geom_hline(
    mapping = aes(
      color = "Michelson's Velocity of Light", 
      yintercept = LIGHTSPEED_MICHELSON
      ),
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    mapping = aes(
      color = "Modern Velocity of Light", 
      yintercept = LIGHTSPEED_VACUUM
      ),
    linetype = "dotted",
  ) +
  geom_point(
    data = . %>% 
      group_by(Distinctness) %>% 
      summarize(VelocityMean = mean(VelocityVacuum)),
    mapping = aes(y = VelocityMean, color = "Mean Velocity")
  ) +
  scale_color_manual(
    values = c(
      "Mean Velocity" = "red",
      "Modern Velocity of Light" = "blue",
      "Michelson's Velocity of Light" = "brown"
    )
  ) +
  labs(
    title = "Michelson's Measurements vs Distinctness",
    y = "Experimental Velocity of Light (km/s)",
    x = "Distinctness"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
    )
```

```{r}
df_q2 %>%
  mutate(
    VelocityUpper = VelocityVacuum + LIGHTSPEED_PM - LIGHTSPEED_VACUUM <= 0,
    VelocityLower = VelocityVacuum - LIGHTSPEED_PM - LIGHTSPEED_VACUUM <= 0
    ) %>%
  mutate(in_bound = VelocityUpper | VelocityLower) %>%
  ggplot() + 
  geom_bar(aes(in_bound, fill = factor(Temp))) +
  scale_fill_grey(start = 0, end = .9) + 
  facet_wrap(
    ~ Distinctness, labeller = as_labeller(
      c(
        "1" = "Distinctness: 1", 
        "2" = "Distinctness: 2", 
        "3" = "Distinctness: 3"
        )
      )
    ) +
  labs(
    fill = "Temperature (°F)",
    x = "Measurement within Uncertainty",
    y = "Count",
    caption = str_wrap("The plot shows the count of measurements that include 
                       the modern speed of light in their uncertainty range.",
                       80)
  ) +
  theme(plot.caption = element_text(hjust = 0.5))
```


**Observations**:

- There seems to be a weak positive correlation between the temperature and the velocity measured
- Measurements with 'Distinctness' 3 has the tightness range of points, and the points that of 1 are more spread out.
- Mean velocities of 'Distinctness' 2 and 3 are very close to LIGHT_MICHELSON
- There are equal number of measurements that are within the uncertainty proposed in each group of 'Distinctness'
- Temperature doesn't seem to have a strong effect on whether the measurements have the modern speed of light in their bound or not.


## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
- [3] BYU video about a [Fizeau-Foucault apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to what Michelson used.
