---
title: "Gapminder"
author: "Daeyoung Kim"
date: 2023-02-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
library(rstatix)
library(ggrepel)
library(latex2exp)
```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
gapminder %>% 
  head(5)
```

**Observations**:

- Write all variable names here
`country`, `continent`, `year`, `lifeExp`, `pop`, and `gdpPercap` are variable in the dataset.

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(pull(gapminder, year))
year_min <- min(pull(gapminder, year))

gapminder %>% 
  filter(year == year_max) %>% 
  filter(continent == "Oceania")
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  mutate(outlier = ifelse(is_outlier(gdpPercap, coef = 1.5), TRUE, FALSE)) %>% 
  ggplot(mapping = aes(continent, gdpPercap)) +
  geom_boxplot(outlier.color = 'red') + 
  scale_y_log10() +
  geom_label_repel(
    data = . %>% filter(outlier == TRUE),
    aes(label = country),
    box.padding = 0.2,
    na.rm = TRUE,
    size = 3
    ) +
  labs( 
    y = "GDP per Capita ($)",
    x = "Continent",
    title = "GDP per Capita vs Continent in 1952"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5)
  )

gapminder %>% 
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  mutate(outlier = ifelse(is_outlier(gdpPercap, coef = 1.5), TRUE, FALSE)) %>% 
  ggplot(mapping = aes(continent, gdpPercap)) +
  geom_boxplot(outlier.color = 'red') + 
  coord_cartesian(ylim = c(200, 15000)) +
  geom_label_repel(
    data = . %>% filter(outlier == TRUE),
    aes(label = country),
    box.padding = 0.2,
    na.rm = TRUE,
    size = 3
    ) +
  geom_text(x = 3, y = 14500, size = 2.5, aes(label = "($108,382)"), data = data.frame(year = 1952)) +
  labs( 
    y = "GDP per Capita ($)",
    x = "Continent",
    title = "GDP per Capita vs Continent in 1952"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5)
  )
```
**Observations**:

- There are significant differences in the median `gdpPercap` with regards to `continent`. 
- There are some countries that earn much more than other countries in the same `continent`. 
- There are very few number of countries that belong to the continent of `Oceania`.
- There are outliers that are above the 75th quantile but not below the 25th quantile. 

**Difficulties & Approaches**:

- Because there is a huge outlier in the continent of Asia, it was extremely difficult to make out much meaningful observations. Therefore, a logarithmic scale was applied to the y-axis to better visualize the plot.
- To have a better understanding of the context of the outliers, I added names of the country as label to the outliers. 
- However, due to the logarithmic scale, it was difficult to observe the relative differences between `continent` that had `gdpPercap` in a different location in the y-axis. Thus, I zoomed in to the plot excluding `Kuwait` with a linear y-axis scale.


### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>% 
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  mutate(outlier = ifelse(is_outlier(gdpPercap, coef = 1.5), TRUE, FALSE)) %>% 
  filter(outlier == TRUE)
```

**Observations**:

- Identify the outlier countries from q2
  - Angola, Bahrain, Canada, Gabon, Kuwait, South Africa, Switzerland, United States, Venezuela.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year == year_max | year == year_min) %>% 
  group_by(year, continent) %>% 
  mutate(outlier = ifelse(is_outlier(gdpPercap, coef = 1.5), TRUE, FALSE)) %>% 
  ggplot(mapping = aes(continent, gdpPercap)) +
  facet_wrap(~year, nrow = 1) +
  geom_boxplot(outlier.color = "red") + 
  # scale_y_log10() +
  coord_cartesian(ylim = c(100, 50000)) +
  geom_label_repel(
    data = . %>% filter(outlier == TRUE),
    aes(label = country),
    box.padding = 0.2,
    na.rm = TRUE,
    size = 2.5,
    nudge_y = 10000,
    nudge_x = -1
    ) +
  geom_text(x = 2.7, y = 48000, size = 2.2, aes(label = "(Asia, $108,382)"), data = data.frame(year = 1952)) +
  labs( 
    y = "GDP per Capita ($)",
    x = "Continent",
    title = "GDP per Capita vs Continent in 1952 and 2007"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0.5)
  )
```

**Observations**:

- As shown by the increased IQR, the disparity between countries in GDP per Capita has drastically increased. 
- Kuwait is no longer an outlier. 
- The disparity between continents in GDP per Capita has also drastically increased looking at the differences in the median. 
- `United States`, `Canada`, `South Africa`, and `Gabon` are the only countries that have stayed as outliers. 

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1-1}
gapminder %>% 
  group_by(year, continent) %>% 
  summarise(median = median(gdpPercap), .groups = "keep") %>% 
  ggplot() +
  geom_point(aes(x = year, y = median, color = continent)) + 
  geom_line(aes(x = year, y = median, color = continent)) +
  labs(
    x = "Year",
    y = "Median GDP per Capita ($)",
    title = "Median GDP per Capita vs Year",
    caption = "The plot shows the median GDP per Capita of different continents over time.",
    color = "Continent"
  )
```

- The median GDP per Capita in different continents show how the the medians diverge over time. 
- The order of the overlaying lines do not change over time; it's constant in the order of Oceania - Europe - Americas - Asia - Africa from top to bottom.
- To more specifically study the observed divergence of the medians, I made a plot of the variance between the medians of each continent over time. 


```{r q5-task1-2}
df_median_variance <- 
  gapminder %>%
  group_by(year, continent) %>%
  summarise(median = as.numeric(median(gdpPercap, na.rm = TRUE)), .groups = "keep") %>% 
  pivot_wider(names_from = continent, values_from = median) %>% 
  # multiplying by n-1/n to get population variance
  data.frame(variance = apply(.[-1], 1, var) * (ncol(.[-1])-1)/ncol(.[-1]))

df_median_variance %>% 
  ggplot() +
  geom_point(aes(year, variance)) +
  geom_line(aes(year, variance)) +
  labs(
    x = "Year",
    y = TeX("Variance of Median GDP per Capita($\\$^2$)"),
    title = "Variance of Median GDP per Capita vs Year",
    caption = "The plot shows the variance of median GDP per Capita of between each continents over time."
  )
```

- Because of the unit of variance, the quantitative value of variance doesn't seem to convey too much meaning. 
- Similar to the observation in the prior plot, the variance exponentially increases over time. 
- This could signify increasing disparities in GDP per Capita among continents. 

```{r q5-task2-1}
gapminder %>% 
  ggplot() +
  geom_point(mapping = aes(x = lifeExp, y = gdpPercap, color = country), size = 0.8, show.legend = FALSE) +
  labs(
    x = "Life Expectancy (year)",
    y = "GDP per Capita ($)",
    title = "GDP per Capita vs Life Expectancy",
    caption = "The plot shows the life expectancy and GDP per Capita of all years of every country present on the dataset."
  )
```

- Initial assumption was that `lifeExp` and `gdpPercap` would have positive correlation. The assumption came from the logic that generally when a country is more economically flourishing, basic survival needs and medical needs would be better fulfilled.
- To validate my assumptions, I first made a scatter plot of `lifeExp` vs `gdpPercap` and saw a monotonic positive correlation between the two variables. 

``` {r q5-task2-2}
df_life_gdp <- 
  gapminder %>% 
  group_by(country, continent) %>% 
  summarize(correlation = cor(lifeExp, gdpPercap,
    method = "spearman",
    use = "na.or.complete"
    ), .groups = "keep"
  )

df_life_gdp %>% 
  ggplot() +
  geom_point(mapping = aes(x = country, y = correlation, color = country), show.legend = FALSE) +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank()) +
  geom_label_repel(
    data = . %>% 
      filter(correlation < 0), 
    aes(x = country, y = correlation, label = country)
    ) +
  labs(
    x = "Country",
    y = "Spearman Correlation Coefficient",
    title = "Correlation between Life Expectancy and GDP per Capita vs Country",
    caption = str_wrap("The plot shows the correlation between life expectancy and GDP per Capita evaluated through Spearman's rank correlation coefficient. The 18 labeled countries are those with negative coefficient values.", width = 120)
  )
```

- To visualize the correlation, I made a plot of Spearman's coefficient vs country. 
- As expected, a lot of countries had either near +1 or positive correlation coefficient. 
- However, there were 18 countries with negative correlation coefficient.
- I filtered the countries with correlation coefficient values less than 0. Out of 18 countries, 14 were African countries. 


``` {r q5-task2-3}
df_life_gdp_cor <-
  df_life_gdp %>%
  mutate(comparison = case_when(continent == "Africa" ~ "Africa", continent != "Africa" ~ "Rest of the World"))

df_life_gdp_cor %>% 
  ggplot() +
  geom_density(mapping = aes(correlation, fill = comparison), alpha=.5) +
  labs(
    fill = "Continent"
  ) +
  labs(
    x = "Spearman Correlation Coefficient",
    y = "Density",
    title = "Density Plot of Spearman Correlation Coefficient",
    caption = str_wrap("The density plot shows the spread of correlation coefficients for life expectancy and GDP per Capita. The plot visualizes the spread for Africa and the rest of the world separately for comparison.", width = 90)
  )
```

- Then to visualize the difference between the spread of the Spearman's correlation coefficient for Africa and the rest of the world, I made a density plot of the correlation coefficient. 
- The spread of the rest of the world was much concentrated near +1, validating my initial assumption for positive correlation. 
- However, Africa's spread covered from positive to negative correlation more evenly.
- As for the *negative* correlation coefficient, I took a step further to investigate into what those might mean contextually. 


```{r q5-task2-4}
gapminder %>% 
  group_by(country) %>% 
  mutate(correlation = cor(lifeExp, gdpPercap,
    method = "spearman",
    use = "na.or.complete"
    )
  ) %>% 
  filter(correlation < 0) %>% 
  filter(continent == "Africa") %>% 
  ggplot() +
  geom_point(mapping = aes(x = lifeExp, y = gdpPercap, color = country)) +
  geom_line(mapping = aes(x = lifeExp, y = gdpPercap, color = country)) +
  geom_label_repel(
    data = . %>% group_by(country) %>% filter(gdpPercap == max(gdpPercap)) %>% filter(gdpPercap > 3000),
    aes(x = lifeExp, y = gdpPercap, label = country)
  ) +
  labs(
    x = "Life Expectancy (year)",
    y = "GDP per Capita ($)",
    title = str_wrap("GDP per Capita vs Life Expectancy for African Countries with Correlation Coefficient < 0", width = 50),
    caption = str_wrap("The plot shows the relationship between life expectancy and GDP per Capita for African countries with negative correlation coefficient values.", width = 90)
  )
```

- To gain a bit more insight into what the actual magnitude of `lifeExp` and `gdpPercap` might be, I went back to the scatter plot to plot only the filtered countries. Except Djibouti and Angola, all the countries had `gdpPercap` less than \$2,000 and `lifeExp` less than about 65 year old. Most countries seemed to have a big range in the `lifeExp`. 
- Here were some research on few of the countries on the plot:
  - Senegal: Long history of separatist war and political instability since the 1980s to present.
  - Mozambique: War of independence since the 1960s to 1975. Series of political turmoil and changes in the political system.
  - Comoros: Independence in the early 1970s. Many political leaders/dictators killed in a coup or assassinations and secessionist movements. Violence on the islands continues even to date. 
- Final thoughts: 
  - Most countries had a positive monotonic correlation between `lifeExp` and `gdpPercap`. 
  - The vast majority of countries with a negative correlation were in Africa. 
  - The GDP per capita of those countries were very low ranging from about \$500 to \$2,000, which makes sense because the median GDP per capita of Africa was much lower than other continents as shown in `q4`. 
  - Many of the countries with negative correlation coefficients in Africa had various political and economical issues in the past. 
